---
title: "DiagrammeR with VivaGraphJS"
author: "Kent Russell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiagrammeR with VivaGraphJS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Way back in my first real post on this blog [Week 01 | Easy Flowcharts and Diagrams ~ DiagrammeR](http://www.buildingwidgets.com/blog/2015/1/8/week-01-easy-flowcharts-and-diagrams-diagrammer), we announced [`DiagrammeR`](http://rich-iannone.github.io/DiagrammeR).  Since then, `DiagrammeR`'s author Rich Iannone has amazed me with his efforts to improve, grow, and document (truly something special) the package.  `DiagrammeR` began with  `mermaid`(`mermaid.js`) and then quickly added `grViz`(`viz.js`/`graphviz`) as two methods to both layout and render diagrams specified as text.

Even with all the dedicated efforts by Rich Iannone, there are still many ways `DiagrammeR` can improve, but much of the improvement depends on some determination of the strategic direction for `DiagrammeR`.  What I think Rich would like is a cohesive set of functionality that allows users easy access to multiple layout engines and interactive renderers while also leveraging strengths of existing R functionality.  With this in mind, areas of improvement can be grouped in three ways:

1. interoperability within `DiagrammeR` - `mermaid` is completely separate and isolated from `grViz` with their own unique specification.
1. interoperability with `R` - `R` is loaded with packages that could benefit `DiagrammeR`.  One example is networks.  Diagrams aren’t always networks, but very often networks become diagrams.  `DiagrammeR`'s support of networks in non-text form, such as `igraph`, does not exist.
1. interactivity - Technically, a user can achieve interactivity in `DiagrammeR`, but only if the user has a thorough understanding of JavaScript.  Incorporating ideas and features like those in the new `htmlwidget` package [visNetwork](http://dataknowledge.github.io/visNetwork/) would provide a very robust and interactive aesthetic result.

## Does VivaGraphJS fit in?

Does the very nice JavaScript library [`VivaGraphJS`](https://github.com/anvaka/VivaGraphJS)/[`ngraph`](https://github.com/anvaka/ngraph) from Andrei Kashcha fit in `DiagrammeR`?  Adding [`VivaGraphJS`](https://github.com/anvaka/VivaGraphJS) gives `DiagrammeR` a interactive rendering engine focused on networks.  If possible, combining `VivaGraphJS` and `grViz` could prove very powerful and synergystic.  Similarly, `VivaGraphJS` could offer nice functionality to `igraph` and other `R` network packages.

I should note that Keeghan Hines already started an `htmlwidget` implementation [`vivagRaph`](https://github.com/keeganhines/vivagRaph) that encouraged me to explore this further.  Thanks so much to Keeghan for his work here.

## Seed for Iteration

Since learning often occurs best by doing, we'll start with a new function `vivagraph` as our seed for iteration.  Hopefully, it will give us some insight into new features, better coordination, additional interactivity, more network support, leveraging of existing R layouts, and additional unforeseen use cases.  Let's see some examples with the first rough draft of `vivagraph`.

## Installation

We are not entirely sure how this will be incorporated, so currently `vivagraph` exists in its own Github branch.  To test out `vivagraph`, please install as shown below.  *Note, existing functionality in `DiagrammeR` will be unaffected, so nothing should break.*

```{r eval=F}
devtools::install_github( "timelyportfolio/DiagrammeR@feature/vivagraphjs" )
```

## With `create_graph` from DiagrammeR

Rich Iannone has put quite a bit of effort into graph building functions that don't require knowledge of the specialized field of networks.  Here is an example using `create_graph` to build our nodes and edges but rendering with `vivagraph`.

```{r}
## please see install step above

library( DiagrammeR )

## use example create_graph from DiagrammeR documentation
##   http://rich-iannone.github.io/DiagrammeR/graphs.html

###
# Create a graph
###

nodes <-
  create_nodes(nodes = LETTERS,
               type = "letter",
               shape = sample(c("circle", "rectangle"),
                              length(LETTERS),
                              replace = TRUE),
               fillcolor = sample(c("aqua", "gray80",
                                    "pink", "lightgreen",
                                    "azure", "yellow"),
                                  length(LETTERS),
                                  replace = TRUE))

edges <-
  create_edges(edge_from = sample(LETTERS, replace = TRUE),
               edge_to = sample(LETTERS, replace = TRUE),
               relationship = "letter_to_letter")


graph <-
  create_graph(nodes_df = nodes,
               edges_df = edges,
               graph_attrs = "layout = neato",
               node_attrs = c("fontname = Helvetica",
                              "style = filled"),
               edge_attrs = c("color = gray20",
                              "arrowsize = 0.5"))

vivagraph( nodes, edges )
```

You might notice some interactivity that we don't get with `grViz`.  However, you will also probably notice that `vivagraph` is not aware of the styling, etc. that we applied on our nodes and edges.  This is definitely on the roadmap for future development.

## With `igraph` network

`DiagrammeR` does not have any built-in handlers for `igraph` and other powerful `R` network packages.  `vivagraph` does try to smartly handle `igraph` network graphs.

```{r}
library( DiagrammeR )
library( igraph )

# plot one of the graph.famous from igraph
vivagraph( graph.famous( "Franklin" ) )
```

You might notice as above that any styling, etc. is not currently supported, but as above, this is definitely on the roadmap.

If you love `igraph` layouts though, you should definitely like this.  In the examples above we used the default `forceDirected` layout from `VivaGraphJS`, but `vivagraph` can also use positions defined by `layout.*`.

```{r}
vivagraph( graph.famous( "Franklin" ), layout = layout.circle )
```

And here is an `igraph` plot for reference using `layout.circle`, so `vivagraph` leverages the layout engine while also adding interactivity and in the future various node types and styling.

```{r}
plot( graph.famous( "Franklin" ), layout = layout.circle )
```

Let's see a couple more to illustrate.

```{r}
vivagraph( graph.famous( "Dodecahedron" ), layout = layout.grid )
```

```{r}
vivagraph( graph.famous( "Herschel" ), layout = layout.fruchterman.reingold )
```

## grViz with igraph layout

As an example how we learn by doing, I discovered that `grViz` could also use `igraph` for layouts.  Even though there is no `vivagraph` in here, let's look at how we could eventually realize interoperability within `DiagrammeR`.  Our diagram will get some additional help from graphviz `splines` and `overlap`.  You can see this at work by closely comparing the bottom of our `grViz` diagram with the reference `igraph` plot below.

```{r}
library(DiagrammeR)
library(igraph)
library(pipeR)

#### use igraph example from documentation
actors <- data.frame(
  name=c("Alice", "Bob", "Cecil", "David", "Esmeralda")
  , age=c(48,33,45,34,21)
  , gender=c("F","M","F","M","F")
)
relations <- data.frame(
  from=c("Bob", "Cecil", "Cecil", "David", "David", "Esmeralda")
  , to=c("Alice", "Bob", "Alice", "Alice", "Bob", "Alice")
  , same.dept=c(FALSE,FALSE,TRUE,FALSE,FALSE,TRUE)
  , friendship=c(4,5,5,2,1,1), advice=c(4,5,5,4,2,3)
)
g <- graph.data.frame(relations, directed=TRUE, vertices=actors)

## use igraph layout for positions
g_layout <- layout.grid(g)

## now plot with DiagrammeR using igraph layout positions
##  but use graphviz splines and overlap handling
g %>>%
  get.data.frame( what = "both" ) %>>%
  (
    # DiagrammeR create_graph helper function
    create_graph(
      data.frame(
        nodes = rownames(.$vertices)
        ,.$vertices
        ,x = g_layout[,1]  # define our x from the igraph layout
        ,y = g_layout[,2]  # define our y from the igraph layout
      )
      ,.$edges
      ,graph_attrs = 'splines="true" overlap = "scale" label = "igraph + grViz"'
    )
  ) %>>%
  (
    grViz( .$dot_code, engine="neato" )
  )
```

Let's compare to the default `igraph` plot.

```{r}
plot( g, layout = g_layout)
```

## Thoughts and Conclusion

Even if `VivaGraphJS` does not fit in `DiagrammeR`, `DiagrammeR` can improve by learning from a `VivaGraphJS` experimental implementation.  **Please give us thoughts, feedback, use cases** to help us determine the future direction of `DiagrammeR`.


## Thanks

Thanks so much

- [`VivaGraphJS`](https://github.com/anvaka/VivaGraphJS) from Andrei Kashcha
- Keeghan Hines for his initial implementation [vivagRaph](https://github.com/keeganhines/vivagRaph)
- Rich Iannone for his dedicated and admirable work on [`DiagrammeR`](http://rich-iannone.github.io/DiagrammeR)
- Ramnath Vaidyanathan and RStudio for `htmlwidgets`
- all the contributors to `R` and `JavaScript`